---
title: "Quality control for the target decoy approach for peptide identification"
author: 
  - name: Lieven Clement
    affiliation:
    - Ghent University
output: 
    html_document:
      code_folding: hide
      code_download: true
      theme: flatly
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
linkcolor: blue
urlcolor: blue
citecolor: blue
---


# load libraries

```{r}
suppressPackageStartupMessages({
library(tidyverse)
library(TargetDecoy)
library(RCurl)
})
```

# Download data from Zenodo
 
```{r}
# Download data from zenodo
options(timeout=300)
url <- "https://zenodo.org/record/7308022/files/search-results.zip?download=1"
destFile <- "searchResults.zip"
if (!file.exists(destFile)) download.file(url, destFile)
unzip(destFile, exdir = "./data", overwrite = TRUE)
```

# Pyrococcus 

## Import Pyrococcus Data in R

```{r}
allTsvFiles <- list.files(path="data",pattern=".tsv$",full.names = TRUE)
msgfFiles <- allTsvFiles[grepl("msgf",allTsvFiles)&grepl("PXD001077",allTsvFiles)]
dfsmsgf <- lapply(msgfFiles, read_tsv)
crapEntries <- scan("db-gpm-crap-entries.txt", what = "character")
dfsmsgf <- lapply(dfsmsgf, function(db) 
  db[rowSums(sapply(crapEntries, grepl,fixed=TRUE,x = db$protein_list)) == 0,])

tandemFiles <- allTsvFiles[grepl("xtandem",allTsvFiles)&grepl("PXD001077",allTsvFiles)]
dfsTandem <- lapply(tandemFiles, read_tsv)
dfsTandem <- lapply(dfsTandem, function(db) 
  db[rowSums(sapply(crapEntries, grepl,fixed=TRUE,x = db$protein_list)) == 0,])
```

## FDR function 

Calculate FDR on tibble with search results. It is assumed that higher scores are better. 

```{r}
# FDR function
fdrDb <- function(db)
{
  db <- db %>% 
    arrange(desc(score)) %>% 
    mutate(FP = cumsum(is_decoy),
           FDR = cumsum(is_decoy)/cumsum(!is_decoy))
  FDR <- db$FDR
  FDRmin <- FDR[length(FDR)]
  for (j in (length(FDR)-1):1)
  {
    if (FDR[j] < FDRmin)
      FDRmin <- FDR[j] else
      FDR[j] <- FDRmin   
  }
  db$FDR <- FDR
  db <- db %>% 
    arrange(spectrum_id) 
  return(db)
}
```

## Generate plots for MSGF+  

Histograms and P-P plots for rank1, rank 2 and search against human DB.

```{r}
# Generate MSGF+ plots pyrococcus
histsMsgf <- lapply(dfsmsgf,function(db, score, decoy, log10) evalTargetDecoysHist(db %>% filter(rank==1), decoy, score, log10) + xlab("Score") + ggtitle(NULL) +   geom_histogram(bins=50, position="identity",alpha=.9), score="score", decoy="is_decoy", log10=TRUE)
histsMsgfR2 <- lapply(dfsmsgf,function(db, score, decoy, log10) evalTargetDecoysHist(db %>% filter(rank==2), decoy, score, log10) + xlab("Score") + ggtitle(NULL) +  geom_histogram(bins=50, position="identity",alpha=.9)   
, score="score", decoy="is_decoy", log10=TRUE)
ppPlotsMsgf <- lapply(dfsmsgf,function(db, score, decoy, log10) evalTargetDecoysPPPlot(db %>% filter(rank==1), decoy, score, log10) + ggtitle(NULL) +xlab("Fd")+ylab("Ft"), score="score", decoy="is_decoy", log10=TRUE)

pyroId <- which(
  grepl(pattern="swissprot",msgfFiles,fixed = TRUE) & grepl(pattern="pfuriosus",msgfFiles,fixed = TRUE)
  ) 

dfHlp <- dfsmsgf[[pyroId]]  %>% 
  filter(rank==1) %>% 
  mutate(score = -log10(score)) %>%
  fdrDb()
thresh <- dfHlp %>%  
  filter(FDR < 0.01) %>% 
  pull(score) %>% 
  min
nSig <- dfHlp %>% filter(FDR < 0.01) %>% pull(is_decoy) %>% `!` %>% sum

humanId <- which(
  grepl(pattern="swissprot",msgfFiles,fixed = TRUE) & grepl(pattern="hsapiens",msgfFiles,fixed = TRUE)
  ) 
ggsave( 
  "./figs/figMsgfSwissHistsPyroR1R2_Human.png", 
  plot =  
    gridExtra::grid.arrange(
      histsMsgf[[pyroId]] +
        geom_histogram(
          bins=50, 
          position="identity",
          alpha=.9) +
         ggtitle(paste(nSig,"target PSMs at 1% FDR")) +
        annotate("rect", xmin = thresh, xmax = 35, ymin = -10, ymax = 1600, alpha = .2) + 
      annotate(geom="text", x=thresh+1, y=1500, label="x > t['1% FDR']",
              color="black",hjust=0 ,parse=TRUE) +
        annotate(geom="rect",xmin=thresh,xmax=thresh,ymin=0,ymax=1600,color="red")+
         theme(legend.position = c(0.75,0.75)),
    histsMsgfR2[[pyroId]] + 
         geom_histogram(
           bins=50, 
           position="identity",
           alpha=.9) +
         ggtitle("Rank 2 PSMs") +
         theme(legend.position = c(0.75,0.75)),
    histsMsgf[[humanId]] + 
         geom_histogram(
           bins=50, 
           position="identity",
           alpha=.9) +
         ggtitle("Spectra matched to H. sapiens") +
         theme(legend.position = c(0.75,0.75)),
    ncol=3),
  device = "png", 
  width = 11.7,
  height = 3.9)

ggsave( 
  "./figs/figMsgfSwissPPplotsPyro_Human.png", 
  plot =  
    gridExtra::grid.arrange(
      ppPlotsMsgf[[humanId]] + 
         ggtitle(NULL),
       ppPlotsMsgf[[pyroId]] + 
         ggtitle(NULL),
       ncol=2),
  device = "png", 
  width = 7.8, 
  height = 3.9)

```

## X-tandem searches


### Preprocess tandem data

Convert spectrum id into double for sorting + add FDR

```{r}
# Preprocess Tandem pyrococcus results
for (i in 1:length(dfsTandem))
{
  
  dfsTandem[[i]] <- dfsTandem[[i]] %>%
    mutate(spectrum_id_orig = spectrum_id,
           spectrum_id = sapply(
             spectrum_id %>% strsplit(split=" "), 
             function(x) substr(x[3],6,1000)) %>% as.double
    )
}
dfsTandem <- lapply(dfsTandem, fdrDb)
```

### Plots for search with and without refinement. 

```{r}
# plots xtandem Results
histsTandem <- lapply(dfsTandem,evalTargetDecoysHist, score="score", decoy="is_decoy", log10=FALSE)
ppPlotsTandem <- lapply(dfsTandem,evalTargetDecoysPPPlot, score="score", decoy="is_decoy", log10=FALSE)

noRefineId <- which(grepl("no-refine",tandemFiles))
refineId <- which(!grepl("no-refine",tandemFiles))

ggsave( 
  "./figs/figTandemRefineSwissHistPP.png", 
  plot =  gridExtra::grid.arrange(
       histsTandem[[refineId]] + 
         geom_histogram(bins=50, position="identity",alpha=.9) +
         ylim(0,1500) +
         xlim(-6.2,35) +
         ggtitle(NULL) +
         theme(legend.position = c(0.75,0.75)),
       ppPlotsTandem[[refineId]] + 
         ggtitle(NULL) +
         ylim(0,.6), 
       ncol=2),
  device = "png", 
  width = 7.8, 
  height = 3.9)
ggsave(
  "./figs/figTandemNoRefineSwissHistPP.png", 
  plot = gridExtra::grid.arrange(
       histsTandem[[noRefineId]] + 
        geom_histogram(bins=50, position="identity",alpha=.9) +
         ylim(0,1500) + 
         xlim(-6.2,35) +
         ggtitle(NULL) + 
         theme(legend.position = c(0.75,0.75)),
       ppPlotsTandem[[noRefineId]] + 
         ggtitle(NULL) +
         ylim(0,.6), 
       ncol=2),
  device = "png", 
  width = 7.8, 
  height = 3.9)
```

### Plot to stratify refinement searched peptides

Calculate 1% FDR threshold for plot 

```{r}
thresh <- dfsTandem[[refineId]] %>%  
  filter(FDR < 0.01) %>% 
  pull(score) %>% 
  min
```

Identify type of PSM: same match, swiched match or new spectrum upon refinement. 

```{r}
dfsTandem[[refineId]]$refine <- "switched"
dfsTandem[[refineId]]$inNoRefine <- dfsTandem[[refineId]]$spectrum_id %in% dfsTandem[[noRefineId]]$spectrum_id

dfsTandem[[refineId]]$refine[!dfsTandem[[refineId]]$inNoRefine] <- "new"

dfRefineCommon <- dfsTandem[[refineId]] %>% filter(inNoRefine)

dfsTandem[[noRefineId]]$inRefine <- dfsTandem[[noRefineId]]$spectrum_id %in% dfsTandem[[refineId]]$spectrum_id
dfNoRefineCommon <- dfsTandem[[noRefineId]] %>% filter(inRefine)
dfRefineCommon$refine[dfRefineCommon$peptidoform ==dfNoRefineCommon$peptidoform] <-"same"

dfsTandem[[refineId]]$refine[dfsTandem[[refineId]]$inNoRefine] <- dfRefineCommon$refine
dfsTandem[[refineId]]$refineTD <- paste(dfsTandem[[refineId]]$refine,ifelse(dfsTandem[[refineId]]$is_decoy,"decoy","target")) %>% factor(levels=c("same target","same decoy","new target","switched target","new decoy","switched decoy"))
```

Construct figure with histograms stratefied according to PSM type. 

```{r}
figsTandemSwissRefinementHistPsmType <- dfsTandem[[refineId]] %>% 
  ggplot(aes(x=score,fill=refineTD, col = I("black"))) +
  geom_histogram(bins=50, position="identity",alpha=.9) +   
  scale_fill_manual("values" = c(`same target` = "#009900",`same decoy` = "#FF9900", `new target` = "#5df542", `new decoy`="#c03600",`switched target`="#42f5c8",`switched decoy` = "#f54281")) +
  theme_bw() + 
  theme(plot.title = element_text(size = rel(1.5)), 
        axis.title = element_text(size = rel(1.2)), 
        axis.text = element_text(size = rel(1.2)), 
        axis.title.y = element_text(angle = 0))

psmType <- dfsTandem[[refineId]] %>% pull(refineTD) %>% table
pi0 <- sum(dfsTandem[[refineId]]$is_decoy)/ sum(!dfsTandem[[refineId]]$is_decoy)
piSame <- psmType["same decoy"]/sum(!dfsTandem[[refineId]]$is_decoy)
piSameNew <- (psmType["same decoy"]+psmType["new decoy"])/sum(!dfsTandem[[refineId]]$is_decoy)
piSameNewSwitch <- (psmType["same decoy"]+psmType["new decoy"] + psmType["switched decoy"])/sum(!dfsTandem[[refineId]]$is_decoy)
piSameNewSwitchNewTarget <- (psmType["same decoy"]+psmType["new decoy"] + psmType["switched decoy"] + psmType["new target"])/sum(!dfsTandem[[refineId]]$is_decoy)

ggsave("./figs/figTandemRefineSwissPSMtypeHistPP.png", 
       plot= gridExtra::grid.arrange(
         figsTandemSwissRefinementHistPsmType +
           ggtitle(NULL) +
           ylab(NULL) +
           xlim(-6.2,35) +
           theme(
             legend.position = c(0.73,0.68),
             legend.title = element_blank()) +
                    annotate("rect", xmin = thresh, xmax = 35, ymin = -10, ymax = 1500, alpha = .2) + 
      annotate(geom="text", x=thresh+1, y=1400, label="x > t['1% FDR']",
              color="black",hjust=0 ,parse=TRUE) +
        annotate(geom="rect",xmin=thresh,xmax=thresh,ymin=0,ymax=1500,color="red"), 
         ppPlotsTandem[[refineId]] + 
           geom_polygon(aes(x=x,y=y),data=data.frame(x=c(0,1,1),y=c(0,0,piSame)),fill="#FF9900",inherit.aes = FALSE) +
           geom_polygon(aes(x=x,y=y),data=data.frame(x=c(0,1,1),y=c(0,piSame,piSameNew)),fill="#c03600",inherit.aes = FALSE) +
           geom_polygon(aes(x=x,y=y),data=data.frame(x=c(0,1,1),y=c(0,piSameNew,piSameNewSwitch)),fill="#f54281",inherit.aes = FALSE) +
           geom_polygon(aes(x=x,y=y),data=data.frame(x=c(0,1,1),y=c(0,piSameNewSwitch,piSameNewSwitchNewTarget)),fill="#5df542",inherit.aes = FALSE) +
           geom_abline(
             slope = pi0) + 
           geom_abline(
             slope = piSameNewSwitchNewTarget) + 
           ylim(0,0.6) + 
           ggtitle(NULL), 
         ncol=2), 
       device = "png", 
       width = 7.8, 
  height = 3.9)
```

# Human sample 

Make diagnostic plots for rank 1 and rank 2. 

```{r}
# P-P plot for human sample 
path2File <- "./data/PXD028735-LFQ_Orbitrap_DDA_Human_01_uniprot-hsapiens-canonical-isoforms-crap_msgfplus-phospho.msgf.mzid.tsv"
df <- read_tsv(path2File)
df <- df[rowSums(sapply(crapEntries, grepl,fixed=TRUE,x = df$protein_list)) == 0,]


ggsave(
  filename = "./figs/figHumanMsgfPlus.png",
  plot = gridExtra::grid.arrange(
    evalTargetDecoysHist(
      df %>% filter(rank==1),
      "is_decoy",
      "score",
      TRUE) + 
      ggtitle(NULL) +
      geom_histogram(
        bins=50, 
        alpha=.9, 
        position="identity") +
      theme(
             legend.position = c(0.75,0.75),
             ),
    evalTargetDecoysPPPlot(
      df %>% filter(rank==1),
      "is_decoy", 
      "score",
      TRUE) + 
      ggtitle(NULL),
    ncol=2
  ),
  device = "png", 
  width = 7.8, 
  height = 3.9)

ggsave(
  filename = "./figs/figHumanMsgfPlusR2.png",
  plot = gridExtra::grid.arrange(
     evalTargetDecoysHist(
      df %>% filter(rank==2),
      "is_decoy",
      "score",
      TRUE) + 
      ggtitle(NULL) +
      geom_histogram(
        bins=50, 
        alpha=.9,
        position = "identity") +
      theme(
             legend.position = c(0.75,0.75),
             ),
    evalTargetDecoysPPPlot(
      df %>% filter(rank==2),
      "is_decoy", 
      "score",
      TRUE) + 
      ggtitle(NULL),
    ncol=2
  ),
  device = "png", 
  width = 7.8, 
  height = 3.9)
```

# Immunopeptidomics 

Make diagnostic plots 

```{r}
# plot immunopeptidomics plot
path2File <- "./data/immunopeptidomics_msms_IAA.txt"
df <- read_tsv(path2File)

ggsave(
  filename = "./figs/figPeptidomics.png",
  plot = gridExtra::grid.arrange(
    evalTargetDecoysHist(df,"is_decoy", "score",FALSE) + 
      geom_histogram(
        bins=50, 
        alpha=.9, 
        position="identity") +
      ggtitle(NULL) +
      theme(
             legend.position = c(0.3,0.7),
             ),
    evalTargetDecoysPPPlot(df,"is_decoy", "score",FALSE) + 
      xlab("Fd") + 
      ylab("Ft") +
      ggtitle(NULL),
    ncol=2
  ),
  device = "png", 
  width = 7.8, 
  height = 3.9)
```

# Joint P-Plot 

Put searches of MSGF+ and XTandem with refinement on one plot. 

```{r}
# Joint PP-plot
msgfId <- which(
  grepl(pattern="swissprot",msgfFiles,fixed = TRUE) & grepl(pattern="pfuriosus",msgfFiles,fixed = TRUE)
  ) 

h <- TargetDecoy:::processObjects(
  list(dfsTandem[[refineId]],dfsmsgf[[msgfId]] %>% filter(rank==1)),
  decoy=c("is_decoy","is_decoy"),
  score=c("score","score"),
  log10=c(FALSE,TRUE)
  ) %>% 
  TargetDecoy:::ppScoresData()
ggsave(
  file = "./figs/figTandemMsGfSwissCombindedPP.png",
  plot = h$df %>%
    ggplot(
      aes(Fdp,z,color=id%>%as.factor)) + 
    geom_point() +
    geom_abline(slope=0) + 
    theme_bw() + 
    theme(legend.title = element_blank(),
          legend.position = c(0.2,.85)) +
    ylab("Ft-pi0") +
    scale_color_discrete(labels=c("X!Tandem","MSGF+")) +
    xlab("Fd"),
  device = "png", 
  width = 3.9, 
  height = 3.9)
```

# Clean up files

```{r}
f <- list.files("./data", recursive = TRUE, full.names = TRUE)
file.remove(f)
file.remove(data)
file.remove(destFile)
```

# Session Info

```{r}
sessionInfo()
```




